Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-29 report/MSc_Report.Rnw
.

\chapter*{Abstract}
% TODO: add abstract text.

\chapter*{Acknowledgments}
% TODO: add acknowledgments.

@@

<<'child-chapter05', child='chapter05.Rnw'>>=
@

>>>>

\chapter*{Abstract}
% Checklist (Abstract): [x] integrate supplied summary [x] confirm PDF rebuild

This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF; additive predictor), and copulas (used only for low-dimensional diagnostics, $K\!\le\!3$). All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable. In the configuration studied (additive predictor and monotone CDF smoothing), TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.

On Half-Moon ($n=250$), mean joint negative log-likelihoods (NLL; lower is better) were $1.71$ (TRTF), $1.93$ (TTM-Sep), and $1.54$ (copula). On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps. On MiniBooNE ($K=43$; sum test log-likelihood), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings. These numbers are not strictly comparable but indicate the relative accuracy of this configuration.

Overall, TRTF tends to lead within the separable family at low dimension, while higher-dimensional datasets expose the limits of separable structure. We report robustness checks (ordering), calibration diagnostics, and the numerical safeguards used, and we outline directions toward richer parameterizations within the same evaluation frame.

@@

% References
\bibliographystyle{plainnat}
\bibliography{biblio}
